Higher-Order unification for free!

### 2024-05-03 paper/main.tex
solve todo in simulation-fidelity of mapdeduplication

%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}\xspace}}{O}} % space non va
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{M}}\xspace}{M}}

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\wellb_{\beta \eta}(\foUnifPb)$,
  if $\;\foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\begin{proof}[Proof sketch]
  By the definition of \mapdeduplication, any two occurrencies of the same \Fo{} 
  variable used at different arities are compiled to two
  variables $A^m$ and $C^n$ related by a chain of \linketa.
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
%
  Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  links is shrinked by $q$. Note that if this length is bigger then $n-m$
  the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  Otherwise,
  if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  due to the constraint of the number of lambda by which $C^n$ should start.
%
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
%
  % If $A^m$ is instantiated, 
%
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}
\todo{this proof is still unreadable}
% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

\noindent
Note that shortening the chain of links allows unification to fail if
$X_1$ and $X_2$ are unified with different terms.
\todo{what are these X1 and X2?}

>>>>

%\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{F}_{\!o}\xspace}}{Fo}} % space non va
%\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{H}_o}\xspace}{Ho}}
\newcommand{\Fo}{\texorpdfstring{\ensuremath{\mathcal{O}}\xspace}{O}}
\newcommand{\Ho}{\texorpdfstring{\ensuremath{\mathcal{M}}\xspace}{M}}

@@

\begin{theorem}[Fidelity with \mapdeduplication]
  Given a list of unification problems \foUnifPb, such that 
  $\wellb_{\beta \eta}(\foUnifPb)$,
  if $\ \foUnifPb$ contains the same \Fo{} variable used at different arities,
  then \mapdeduplication guarantees \fullref{prop:fidelity}
\end{theorem}

\def\chain{\ensuremath{\mathcal{C}}\xspace}

\begin{proof}[Proof sketch]
  Let $X$ a \Fo variable appearing in two different subterms $s_1$ and $s_2$. in
  $s_1$, $X$ is applied to the list of terms $x_1\dots x_m$ and, in $s_2$, it is
  applied to $y_1\dots y_n$. Here, $m < n$. By \mapdeduplication, the
  compilation genereates two distinct \Ho variables for $X$, say $A$ and $C$,
  related by a chain \chain of \linketa{}s, with length $n - m$. We prove that
  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
  these terms fails to unify. By \progressetaleft, the assignment of $A$
  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
  chain collapses and unification is performed between $\lambda x_n\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
  we work with \wellb terms, this unification succeeds iff it does in \Fo. If,
  otherwise, $p < n$, then two cases should be considered: $t_1$ is a rigid
  term, then, $t_2$ must be an eta expansion of $t_1$, implying $q = 0$. If it
  is not the case, then the successive progressions made by \progressetaleft
  would eventually cause a failure. If, on the other hand, $t_1$ is a
  unification variable, \elpiIn{scope-check} guarantees that $t_2$ does not use
  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
  decompilation is charged wake all links in \linkStore and eventually relate
  $t_1$ with $t_2$.
  %
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
  % 
  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
  %
  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  % Otherwise,
  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  % due to the constraint of the number of lambda by which $C^n$ should start.
  %
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
  %
  % If $A^m$ is instantiated, 
  %
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}

% \todo{this proof is still unreadable}
% \todo{Not sure I fullfy fixed the proof, but the old oen was not ok}

% Note that shortening the chain of links allows unification to fail if
% $X_1$ and $X_2$ are unified with different terms.
% \todo{what are these X1 and X2?}
